---
title: "UNICEF Project Exploration - STA130 Winter 2024 - Group D"
author: "Brad Cho, Tanay Langhe, Christopher Li, Dhruv Patel, Chaoxun Tan"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Final Project Overview: Identifying Opportunities to Accelerate Progress on Sustainable Development Goals (SDG)

How do disparities in countries’ socio-economic status relate to their progress towards reaching the SDGs?

## How do we measure a country's socio-economic status?

To measure a country’s socio-economic status, we look at its GDP collected by the World Bank in 2022/2021. In order to determine the socioeconomic brackets, we put all of the countries’ gdp onto a log10 scale and use the k-means algorithm to split them into 4 distinct groups: low, medium, high, and very-high.

Cleaning Data
```{r}
# To make sure everything is reproducable
set.seed(69420)

# Cleaning GDP Data
gdp_data <- read_csv("data/country_gdps.csv") %>%
  mutate(log_recent = case_when(!is.na(`2022`) ~ log10(`2022`),
                                .default = log10(`2021`))
                                ) %>%
  mutate(country_code = `Country Code`) %>%
  select(country_code, log_recent) %>%
  filter(!is.na(log_recent))
```

Calculate and remove outliers (1.5 IQR) - This is not used!

```{r}
if (FALSE) {
  median = median(gdp_data$log_recent)
  iqr = IQR(gdp_data$log_recent)

  gdp_data <- gdp_data %>% filter(median - 1.5 * iqr < log_recent & 
                                    log_recent < median + 1.5 * iqr)
}
```

Plotting GDP Data

```{r}
gdp_data %>% ggplot(aes(x = log_recent)) + geom_histogram(bins=20)
```

Elbow Method to figure out the best number of clusters, we determined 3 or 4 clusters is best

```{r}
explained_ss <- rep(NA, 10)
for(k in 1:10){
  # run k-means on the data
  clustering <- kmeans(gdp_data$log_recent, k)
  explained_ss[k] <- clustering$betweenss / clustering$totss
}

ggplot() +
  aes(x=1:10, y=1-explained_ss) +
  geom_line() +
  geom_point() +
  labs(x="Number of Clusters",
       y="Remaining Variation",
       title="K-Means Clustering Performance") +
  theme(text=element_text(size=18)) +
  scale_x_continuous(breaks=1:10) +
  scale_x_continuous(breaks=1:10)
```

Actually performing k-means on the data, lower number group indicate lower gdp group

```{r}
clustering <- kmeans(gdp_data$log_recent, 4)
gdp_data <- gdp_data %>% mutate(cluster = clustering$cluster) %>% 
                         mutate(cluster = case_when(cluster == 3 ~ 2, 
                                                    cluster == 4 ~ 3, 
                                                    cluster == 2 ~ 4, 
                                                    cluster == 1 ~ 1))
```

Plotting k-means

```{r}
gdp_data %>% ggplot(aes(x=log_recent, group=cluster, fill=cluster)) + geom_histogram(bins = 10)
```

Writing cleaned CSV to a file, for the purpose of this rmd file, we will just be setting it to a variable

```{r}
# write.csv(gdp_data, "data/clean_country_gdps.csv", row.names=FALSE)
clean_country_gdps <- gdp_data
```

## Research Question 1

In the second-lowest socio-economic bracket, if we group countries based on location, is the average distance towards the SDG’s significant between the groups?

Google's data-set uses Alpha-2 code, I need to convert to Alpha-3 since UNICEF uses Alpha-3

```{r}
country_locations <- read_csv("data/country_locations.csv") %>% select(-name)
country_codes <- read_csv("data/country_codes.csv") %>% 
    filter(`Alpha-2 Code` %in% country_locations$country) %>% 
    mutate(alpha_2 = `Alpha-2 Code`, alpha_3 = `Alpha-3 Code`) %>% 
    select(-`Alpha-2 Code`, -`Alpha-3 Code`, -Country) 
country_locations <- country_locations %>% filter(country %in% country_codes$alpha_2) %>% 
    mutate(alpha_2 = country) %>% select(-country)

merged_data <- merge(country_locations, country_codes, by="alpha_2") %>% 
    mutate(country_code = alpha_3) %>% select(country_code, latitude, longitude)
```

Writing cleaned CSV to a file, for the purpose of this rmd file, we will just be setting it to a variable

```{r}
# write.csv(merged_data, "data/clean_country_locations.csv", row.names=FALSE)
clean_country_locations <- merged_data
```

### Doing k-means on the location

Cleaning datasets

```{r}
set.seed(69420)

country_locations = data.frame(clean_country_locations)
gdp_data = data.frame(clean_country_gdps) %>% filter(cluster==2 & country_code %in% country_locations$country_code) %>% select(-cluster)

merged_data = merge(gdp_data, country_locations, by="country_code")
```

Elbow Method to figure out the best number of clusters, we determined 3 or 4 clusters is best
```{r}
explained_ss <- rep(NA, 10)
for(k in 1:10){
  # run k-means on the data
  clustering <- kmeans(merged_data %>% select(longitude, latitude), k)
  explained_ss[k] <- clustering$betweenss / clustering$totss
}

ggplot() +
  aes(x=1:10, y=1-explained_ss) +
  geom_line() +
  geom_point() +
  labs(x="Number of Clusters",
       y="Remaining Variation",
       title="K-Means Clustering Performance") +
  theme(text=element_text(size=18)) +
  scale_x_continuous(breaks=1:10) +
  scale_x_continuous(breaks=1:10)
```

Doing k-means with 4 clusters

```{r}
clustering <- kmeans(merged_data %>% select(longitude, latitude), 4)
merged_data <- merged_data %>% mutate(cluster = clustering$cluster)
```

Plotting k-means

```{r}
# TODO - add a map projection behind this?
merged_data %>% ggplot(aes(x = longitude, y = latitude, color=cluster)) + 
                    geom_point() + xlim(-180, 180) + ylim(-90, 90)
```


Writing cleaned CSV to a file, for the purpose of this rmd file, we will just be setting it to a variable

```{r}
# write.csv(merged_data, "data/country_locations_cluster.csv", row.names=FALSE)
country_locations_cluster <- merged_data
```


## Doing hypothesis testing on the different groups

```{r}
country_locations <- data.frame(country_locations_cluster)
sdr_goals <- read_csv("data/sdr_fd5e4b5a.csv") %>% 
    mutate(country_code = `Country Code ISO3`, 
           score = `2023 SDG Index Score`) %>% 
    select(country_code, score) %>% 
    filter(country_code %in% country_locations$country_code)

merged_data <- merge(sdr_goals, country_locations, by="country_code") %>% 
    filter(!is.na(score))
```

Summaries

```{r}
print(merged_data %>% group_by(cluster) %>% summarize(m = mean(score)))

for (i in 1:3) {
  for (j in (i+1):4) {
    print(paste("Group", i, "with Group", j));
    model1 <- lm(
      score ~ cluster, 
      data = merged_data %>% filter(cluster == i | cluster == j) %>% 
        mutate(cluster =  as.character(cluster))
    )
    print(summary(model1)$coefficients)
  }
}
```


Grouped Boxplot visualization

```{r}
merged_data %>% ggplot(aes(x=cluster, y=score, group=cluster, fill=cluster)) + 
                  geom_boxplot()
``` 

## Question 2

For each socio-economic bracket, what is a range of plausible values for the success of education SDG’s?

```{r}
set.seed(694201)

cleangdps <- data.frame(clean_country_gdps)
cleanind <- read_csv("data/Very_clean_country_indicators.csv")

merged_data <- merge(cleangdps, cleanind, by.x = "country_code", by.y = "iso3", 
                     all.x = TRUE)

# merging da data
bootstrap_ci <- function(data, n_bootstrap = 1000) {
  bootstrap_means <- replicate(n_bootstrap, mean(sample(data, replace = TRUE), 
                                                 na.rm = TRUE))
  ci <- quantile(bootstrap_means, probs = c(0.025, 0.975))
  list(mean = mean(bootstrap_means), ci_lower = ci[1], ci_upper = ci[2], 
       bootstrap_means = bootstrap_means)
}
```

Doing bootstrapping

```{r}
# applying the bootstrap Function and preparing the data for plotting
bootstrap_results <- merged_data %>% 
  filter(!is.na(sowc_education__completion_completion_rate_2013_2021_avg)) %>% 
  group_by(cluster) %>%
  summarise(bootstrap_data = list(bootstrap_ci(
      sowc_education__completion_completion_rate_2013_2021_avg
    )),
            .groups = 'drop')
```

Plotting the results

```{r}
# Extracting results for plotting
bootstrap_distributions <- do.call(rbind, lapply(1:nrow(bootstrap_results), function(i) {
  data.frame(cluster = bootstrap_results$cluster[i], 
             bootstrap_means = bootstrap_results$bootstrap_data[[i]]$bootstrap_means)
}))

ci_data <- do.call(rbind, lapply(1:nrow(bootstrap_results), function(i) {
  with(bootstrap_results$bootstrap_data[[i]],
       data.frame(cluster = bootstrap_results$cluster[i], 
                  mean = mean, 
                  ci_lower = ci_lower, 
                  ci_upper = ci_upper))
}))
# ploting bootstrap distributions
ggplot(bootstrap_distributions, aes(x = bootstrap_means, fill = as.factor(cluster))) +
  geom_density(alpha = 0.5) +
  labs(title = "Bootstrap Distributions of Education Completion Rates", 
       x = "Bootstrap Means", 
       y = "Density", 
       fill = "Cluster") +
  theme_minimal()

# plotting for confidence intervals
ggplot(ci_data, aes(x = as.factor(cluster), y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  labs(title = "95% Confidence Intervals for Education Completion Rates by Cluster", 
       x = "Cluster", 
       y = "Mean Education Completion Rate") +
  theme_minimal()
```

## Question 3

How well does one’s socio-economic status correlate with a country’s progress towards the health SDGs?

We are using Goal 1 as a different measure of a country's socio-economic status

```{r}
sdr_goals <- read_csv("data/sdr_fd5e4b5a.csv") %>% 
  mutate(goal_1 = `Goal 1 Score`, goal_3 = `Goal 3 Score`) %>%
  select(goal_1, goal_3) %>% filter(!is.na(goal_1) & !is.na(goal_3))
```

Plotting the result

```{r}
sdr_goals %>% ggplot(aes(x=goal_1, y=goal_3)) + geom_point() + 
                geom_smooth(method="lm", se=FALSE)
```

Model Coefficients

```{r}
model <- lm(goal_3 ~ goal_1, data = sdr_goals)
summary(model)$coefficients
```
Plotting residual plot

```{r}
df <- fortify(model)
df %>% ggplot(aes(x = goal_1, y = .resid)) + geom_point() + geom_smooth(method="lm", 
                                                                        se=FALSE)
```


